\documentclass[10pt,a4paper,final,oneside,openany]{memoir}

\input{defaultprelude.tex}

\usepackage[titelside, en, nat]{ku-forside}

\titel{Exploiting functional invariants}
\undertitel{to optimise parallelism: a dataflow approach}
\forfatter{Troels Henriksen -- \texttt{athas@sigkill.dk}}
\forfatterET{}
\forfatterTO{}
\opgave{Master's Thesis} % Findes kun under 'titelside'
\dato{February 2014}
\vejleder{Supervisors: Cosmin Eugen Oancea and Fritz Henglein}

\newcommand\boolt[0]{\texttt{bool}}
\newcommand\realt[0]{\texttt{real}}
\newcommand\chart[0]{\texttt{char}}
\newcommand\intt[0]{\texttt{int}}
\newcommand\aliases[1]{\textrm{aliases}(#1)}
\renewcommand\tilde[0]{{\raise.17ex\hbox{$\scriptstyle\sim$}}}
\newcommand{\LO}{$\mathcal{L}_0$}

\newcommand{\tagsc}[1]{\tag{\textsc{#1}}} % Fake Small Caps tagging
\newcommand{\nfrac}[2]{\frac{\displaystyle{#1}}{\displaystyle{#2}}}
\newcommand*{\bfrac}[2]{\genfrac{}{}{0pt}{}{#1}{#2}}

\definecolor{DikuRed}{RGB}{130,50,32}
\newcommand{\emp}[1]{\textcolor{DikuRed}{ #1}}
\definecolor{CosGreen}{RGB}{10,100,70}
\newcommand{\emphh}[1]{\textcolor{CosGreen}{ #1}}

\newcommand{\mymath}[1]{$ #1 $}
\newcommand{\myindx}[1]{_{#1}}
\newcommand{\myindu}[1]{^{#1}}
\newcommand{\mymathbb}[1]{\mathbb{#1}}
\newcommand{\cindx}[1]{\(\myindx{#1}\)}

% Write "Part II" etc. in TOC instead of just "II"
\renewcommand*{\cftpartname}{Part\space}

\bibliography{bibliography}

\begin{document}
\frontmatter

 \clearpage

\maketitle
~
\vspace{3cm}
  \begin{abstract}
    We present a pure functional data-parallel programming language
    supporting nested parallelism on regular data.  Our language is
    aimed at execution on massively parallel SIMD hardware, such as
    modern graphics-programming units.  The language contains novel
    features for performing in-place modification of arrays without
    violating functional purity.

    Furthermore, our language includes an assertion system that can be
    used to express dynamic checks such as bounds checks directly in
    the language, potentially facilitating their removal from inner
    loops through standard optimisations.

    We discuss the design and implementation of several optimisations,
    notably an advanced hoister, and loop fusion based on a structural
    transformation, which is capable of fusing operations where the
    output is used in multiple places, when possible without
    duplicating computation.  This is an unusual feature among fusion
    transformations.

    The benefit of our optimisations are demonstrated on three
    real-world benchmarks translated to our programming language.  We
    argue that always refusing to duplicate computation is too
    conservative on parallel hardware, and discuss potential
    directions for further improvement.
  \end{abstract}

% \clearpage
%  ~\thispagestyle{empty}
\clearpage
\tableofcontents*
%\openany
\input{preface.tex}
%\openright
\mainmatter
\counterwithout{table}{chapter}
\counterwithout{figure}{chapter}
\input{intro.tex}

\part{Language Design}
\label{part:languagedesign}

\input{l0language.tex}

\input{uniquenesstypes.tex}

\input{internalrepresentation.tex}

\part{Optimisations}

\input{first-order-optimisations.tex}

\input{fusion.tex}

\input{fusion-enabling-soac-transformations.tex}

\input{hindrance-removal.tex}

\part{Evaluation}

\input{optimisation-results.tex}

\input{conclusions.tex}

\clearpage

\part{Closing Credits}

% We want the bibliography in the ToC, but it shouldn't have a chapter
% number.
\phantomsection
\addcontentsline{toc}{chapter}{Bibliography}
\defbibheading{bibliography}{\chapter*{Bibliography}}
\printbibliography

\backmatter
\appendix
\input{appendix.tex}

\end{document}
